---
title: "Transposable elements (TE) analysis in public datasets spanning preimplantation mouse development"
date: "April 11th, 2023"
author: 
- name: Sara Formichetti
  affiliation: EMBL Rome
  email: sara.formichetti@embl.it
output:
  prettydoc::html_pretty:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: yes
---

```{r Setting general options, include=FALSE}

knitr::opts_chunk$set(autodep = TRUE, cache = TRUE, echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')

```

```{r loading needed packages}

library(ggplot2)
library(data.table)
library(pheatmap)
library(DESeq2)
library(gridExtra)
library(ggrepel)
library(ggpubr)

```

# Reading TE counts

```{r samples table}

# loading samples table
samples_table <- fread("../../data/embryo_public_data_GSE66582_GSE76505/sequencing/samples_table_TE_analysis.csv")
names(samples_table)[grep("group_or_time_point", names(samples_table))] <- "stage"

```

```{r load featureCounts output tables}

# specifying the directory containing featureCounts outputs
dir <- "../../data/embryo_public_data_GSE66582_GSE76505/sequencing/featureCounts/STAR_TE"

# getting file names of featureCounts output count tables, sorted alphabetically
files <- sort(list.files(path = dir)[grep("counts.txt$", list.files(path = dir))])

# character vector of names of files with path
files_w_path <- file.path(dir, files)

# reading the count tables: first line is a commented line containing featureCounts command and needs to be skipped; columns chr,start,end,strand are dropped because they contain the comma-separated list of chr,start,end,strand for all occurrences of the TE with that unique repName in the annotation - in case of single copy genes these columns usually contain one value, in this case a list of them because each element is repeated in the genome.
counts_list <- lapply(files_w_path, function (f) {fread(f, skip = 1, drop = c("Chr", "Start",	"End", "Strand"))})
# all counts tables have the same dimensions because featureCounts counts for all meta-features (and puts 0 in case there are 0 count for that feature in that sample), therefore there will be no NAs in the reduced dt
counts_dt <- Reduce(function(...) merge(..., by = c("Geneid", "Length"), all = TRUE), counts_list)
# remove path from sample names
names(counts_dt) <- sub(".*/", "", sub("_amgm350.TE_only.bam", "", names(counts_dt)))
names(counts_dt)[1] <- "repName"

rm(counts_list)

```

```{r merging counts with TE annotation}

# loading TE annotation
TE_annotation_bed <- "../../data/annotations/RepeatMasker_RepeatLibrary20140131_mm10.noGenes.noSimple.bed"
TE_anno <- fread(TE_annotation_bed, fill = TRUE, col.names = c("seqnames", "start", "end", "repName", "strand", "repFamClass"), drop = 5)
# adding +1 to start is necessary for compatibility with methylBase-derived objects because bed file coordinates are 0-based
TE_anno$start <- TE_anno$start + 1
# making column with repFamilies only and column with repClasses only
TE_anno$repFamily <- gsub(".*/", "", TE_anno$repFamClass)
TE_anno$repClass <- gsub("/.*", "", TE_anno$repFamClass)

# merging the count table with TE annotation to get repFamilies and Classes
counts_dt <- merge(counts_dt, unique(TE_anno[,c("repName", "repClass", "repFamily", "repFamClass")]), by = "repName", all.x = TRUE, all.y = FALSE)
# removing rows whose repFamClass is NA
counts_dt <- counts_dt[!is.na(counts_dt$repFamClass)]

```

The input files are:

* the output tables made by featureCounts (run in rule 'count_on_TE' of the Snakefile), which contain read counts of all repetitive elements (i.e. all repNames) for all samples
* the repetitive elements annotation `r TE_annotation_bed` (converted to SAF format in rule of Snakefile) used for featureCounts command, that is used to retrieve repFamily and repClass for each repName. The annotation I chose is the most updated library to Nov2020 of rmsk, from which I removed 'Simple repeats' and 'Low complexity regions'
* a table with total number of reads per sample in raw fastq files 
* the samples' table, containing metadata (i.e. stage)

```{r ordering based on stage}

# releveling stage for plotting
samples_table$stage <- factor(samples_table$stage, levels = c("MII_oocyte", "zygote", "early_2cell", "2cell", "4cell", "8cell", "E35_ICM", "Inner_cell_mass", "E35_TE", "E40_ICM", "E55_Epi", "E55_VE", "E65_Epi", "E65_VE", "E75_Ecto", "E75_PS", "E75_Meso", "E75_Endo"))
setkey(samples_table, stage)
# ordering counts dt based on samples table
setcolorder(counts_dt, c("repName", "repClass", "repFamily", "repFamClass", samples_table$sample))

```

# Analysis at family level 

## FPKM for each family of Repetitive Elements

For each family of Repetitive Elements (in case of elements with no repFamily name or repFamilies belonging to more than one repClass I use repClass) I compute **FPKM** values, as follows: for each sample:

* I compute the sum of counts for all elements belonging to that repFamily
* I divide this sum by the total number of reads for that sample and multiply by 10â¶
* I divide this number by the total sum of lengths (in Kb) of the elements belonging to that repFamily --> FPKM
* When specified, **I subtract from each FPKM the total FPKM of all transposons belonging to the DNA repClass**

```{r load library size table}

# loading table containing total number of STAR input reads per sample
library_size_dt <- fread("../../data/embryo_public_data_GSE66582_GSE76505/sequencing/stats/STAR_TE/STAR_input_reads.txt")
library_size_dt$sample <- gsub(".*/", "", gsub("_amgm350.*", "", library_size_dt$V1))
library_size_dt$STAR_input <- library_size_dt$V2
library_size_dt <- library_size_dt[, .(sample, STAR_input)]
# since I counted fragments in featureCounts, I divide the STAR input reads by 2 to get a rough estimation of read pairs
library_size_dt$library_size <- library_size_dt$STAR_input/2
# removing bad and poorly injected samples
library_size_dt <- library_size_dt[library_size_dt$sample %in% samples_table$sample, ]
# ordering library size as samples and counts table
library_size_dt <- library_size_dt[match(samples_table$sample, library_size_dt$sample)]

```

```{r family FPKM}

# I notice that some elements do not have a repFamily but they have a repClass and that some repFamilies are assigned to more than one repClass, therefore I use 'repFamClass' to compute the FPKM 
# selecting columns on which to compute the sum by 'repFamClass'
cols <- c("Length", names(counts_dt)[names(counts_dt) %in% samples_table$sample])
# sum by 'repFamClass'
family_counts_dt <- counts_dt[, lapply(.SD, sum), by = .(repFamClass), .SDcols = cols] 
# remove rows containing only 0
family_counts_dt <- family_counts_dt[rowSums(family_counts_dt[, cols[2:length(cols)], with = FALSE]) > 0, ]
# divide each column by library size and multiply by 10^6
family_FPKM_dt <- sweep(family_counts_dt[, cols[2:length(cols)], with = FALSE], 2, library_size_dt$library_size/1000000, FUN = '/')
# divide each row by repFamClass total length
family_FPKM_dt <- as.data.table(sweep(family_FPKM_dt, 1, family_counts_dt$Length/1000, FUN = '/'))
family_FPKM_dt$repFamClass <- family_counts_dt$repFamClass
# subtract DNA transposons' FPKM from each other family's FPKM value
DNA_TE_FPKM <- setNames(as.numeric(colSums(family_FPKM_dt[grep("DNA/", family_FPKM_dt$repFamClass), 1:(ncol(family_FPKM_dt)-1)])), cols[2:length(cols)])
family_FPKM_minusDNA_dt <- as.data.table(sweep(family_FPKM_dt[, 1:(ncol(family_FPKM_dt)-1)], 2, DNA_TE_FPKM, FUN = "-"))

# add back columns with repFamilies and repClasses
family_FPKM_dt$repFamily <- gsub(".*/", "", family_FPKM_dt$repFamClass)
family_FPKM_dt$repClass <- gsub("/.*", "", family_FPKM_dt$repFamClass)
family_FPKM_minusDNA_dt$repFamClass <- family_FPKM_dt$repFamClass
family_FPKM_minusDNA_dt$repFamily <- family_FPKM_dt$repFamily
family_FPKM_minusDNA_dt$repClass <- family_FPKM_dt$repClass

```

## Heatmaps

The heatmaps are scaled by rows.

```{r heatmap function}

# Custom function for heatmap of repetitive elements from data.table of FPKM
all_samples_TE_heatmap <- function (my_measures_table, my_samples_table, sample_cluster, elements_cluster, my_title, scale_col) {
  # making df with only the numeric values
  pheat_df <- as.data.frame(my_measures_table)[, colnames(my_measures_table) %in% my_samples_table$sample]
  # rownames appearing on heatmap will be the repFamily names (which are actually the repClass when there is no repFamily assigned to an element)
  repFamilies <- gsub(".*/", "", my_measures_table$repFamClass)
  # in cases when one repFamily name is associated to more than one repClass, duplicates are created in repFamilies vector. Using repFamClass in these cases
  ambigous_rows <- which(duplicated(repFamilies) | duplicated(repFamilies, fromLast=TRUE))
  repFamilies[ambigous_rows] <- my_measures_table$repFamClass[ambigous_rows]
  rownames(pheat_df) <- repFamilies
  # ordering stages
  my_samples_table$stage <- factor(my_samples_table$stage, levels = c("MII_oocyte", "zygote", "early_2cell", "2cell", "4cell", "8cell", "E35_ICM", "Inner_cell_mass", "E35_TE", "E40_ICM", "E55_Epi", "E55_VE", "E65_Epi", "E65_VE", "E75_Ecto", "E75_PS", "E75_Meso", "E75_Endo"))
  setkey(my_samples_table, stage)
  pheat_df <- pheat_df[, my_samples_table$sample]
  # annotating samples with metadata
  pheat_anno_samples <- data.frame(stage = my_samples_table$stage, row.names = my_samples_table$sample)
  # annotating elements with repClasses
  pheat_anno_elem <- data.frame(repClass = my_measures_table$repClass, repFamilies, row.names = rownames(pheat_df))
  # ordering TE for better visualization in case no elements clustering is required
  pheat_df <- pheat_df[order(pheat_anno_elem$repClass), ]
  pheat_anno_elem <- pheat_anno_elem[rownames(pheat_df), ]
  pheat_anno_elem$repFamilies <- NULL
  # removing rows and columns having only 0s
  pheat_df <- pheat_df[rowSums(pheat_df) != 0, colSums(pheat_df) != 0]
  if (scale_col) {
    rn <- rownames(pheat_df)
    pheat_df <- as.data.frame(apply(pheat_df, 2, scale))
    rownames(pheat_df) <- rn
  }
  # plotting heatmap
  pheatmap(pheat_df, cluster_rows = elements_cluster, cluster_cols = sample_cluster, annotation_col = pheat_anno_samples, annotation_row = pheat_anno_elem, show_rownames = TRUE, clustering_distance_rows = "correlation", clustering_distance_cols = "correlation", fontsize = 12, scale = "row", main = my_title)
}

```

```{r heatmap subset of elements preimpl, include = TRUE, fig.height=12, fig.width=15}

preimpl_samples_table <- samples_table[samples_table$stage %in% c("MII_oocyte", "zygote", "early_2cell", "2cell", "4cell", "8cell", "E35_ICM", "Inner_cell_mass", "E35_TE", "E40_ICM"),]

all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n corrected for DNA traspososons", scale_col = FALSE)

```

Based on above plots, I remove the early 2-cell sample SRR2927028 and replot.

```{r heatmap subset of elements preimpl after filtering, include = TRUE, fig.height=12, fig.width=15}

preimpl_samples_table <- samples_table[samples_table$stage %in% c("MII_oocyte", "zygote", "early_2cell", "2cell", "4cell", "8cell", "E35_ICM", "Inner_cell_mass", "E35_TE", "E40_ICM") & samples_table$sample != "SRR2927028",]

all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n not corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "RNA TE FPKM,\n corrected for DNA traspososons", scale_col = FALSE)
all_samples_TE_heatmap(my_measures_table = family_FPKM_minusDNA_dt[grepl("rRNA|DNA|RC|Satellite", family_FPKM_minusDNA_dt$repFamClass)], my_samples_table = preimpl_samples_table, sample_cluster = FALSE, elements_cluster = FALSE, my_title = "Contaminants FPKM,\n corrected for DNA traspososons", scale_col = FALSE)

```

* Note that when correcting for DNA FPKM preimplantation embryonic stages get penalized compared to the oocyte i.e. oocyte FPKM values are overestimated.
* The only clear changes emerging above noise from the DNAFPKM-corrected heatmaps are the well-known ERVL upregulation at the late 2-cell stage and ERVL-MaLR upregulation from the zygote to the late 2-cell stage.

# DE-Seq analysis of RNA transposons between subsequent preimplantation stages

I include the **FPKM of DNA repetitive elements as confounding factors in DESeq2 formula**.

```{r creating DESeq inputs}

#preimpl_samples_table <- samples_table[samples_table$batch == 1,]
preimpl_samples_table <- samples_table[samples_table$stage %in% c("MII_oocyte", "zygote", "early_2cell", "2cell", "4cell", "8cell", "E35_ICM", "Inner_cell_mass", "E35_TE", "E40_ICM")]
preimpl_counts_dt <- counts_dt[, c("repName", "repClass", "repFamily", "repFamClass", preimpl_samples_table$sample, "Length"), with = FALSE]

# Selecting only RNA transposons and only columns with numeric values
RNA_TE_counts_dt <- preimpl_counts_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", preimpl_counts_dt$repClass), preimpl_samples_table$sample, with = FALSE]

# A count matrix will be the input for DESeq
counts_m <- as.matrix(RNA_TE_counts_dt)
rownames(counts_m) <- preimpl_counts_dt[!grepl("DNA|rRNA|\\?|scRNA|Other|snRNA|tRNA|Unknown|srpRNA|RC|^RNA|Satellite", counts_dt$repClass), ]$repName
# the design df is created from the samples_table
design_df <- data.frame(stage = preimpl_samples_table$stage, batch = preimpl_samples_table$batch)
# the rownames of the design df have to be the same (and in the same order) of the colnames of the count matrix
rownames(design_df) <- preimpl_samples_table$sample
# adding another DNA transposons FPKM as an additional variable, that can be used as counfounding factor
design_df$sample <- rownames(design_df)
design_df$DNA_FPKM <- DNA_TE_FPKM[design_df$sample]
design_df$batch <- as.character(design_df$batch)

```

```{r creating DESeq Dataset}

dds <- DESeqDataSetFromMatrix(countData = counts_m,
                              colData = design_df,
                              design = ~ DNA_FPKM + stage)

```

Before running the Differential Expression analysis, the data are pre-filtered to remove all repetitive elements with < 10 reads among all samples.

```{r prefiltering}

keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

```

```{r DESeq function}

dds <- DESeq(dds)

```

```{r adj pvalue cutoff}

adjpvalue_cutoff <- 0.1
l2fc_cutoff_to_label <- 0.2
baseMean_cutoff_to_label <- 10

```

```{r DESeq results list}

# Function which extracts DESeq results (with shrunken log2FC) for all comparisons
get_results_all_comp <- function (my_dds) {
  res_list <- list(
    zygote = lfcShrink(my_dds, contrast=c("stage","zygote","MII_oocyte"), alpha=adjpvalue_cutoff, type="ashr"),
    postEGA_2c_to_zygote = lfcShrink(my_dds, contrast=c("stage","2cell","zygote"), alpha=adjpvalue_cutoff, type="ashr"),
    fourCell = lfcShrink(my_dds, contrast=c("stage","4cell","2cell"), alpha=adjpvalue_cutoff, type="ashr"),
    eightCell = lfcShrink(my_dds, contrast=c("stage","8cell","4cell"), alpha=adjpvalue_cutoff, type="ashr"),
    ICM = lfcShrink(my_dds, contrast=c("stage","Inner_cell_mass","8cell"), alpha=adjpvalue_cutoff, type="ashr"),
    E35_ICM = lfcShrink(my_dds, contrast=c("stage","E35_ICM","8cell"), alpha=adjpvalue_cutoff, type="ashr"),
    E35_TE = lfcShrink(my_dds, contrast=c("stage","E35_TE","8cell"), alpha=adjpvalue_cutoff, type="ashr"),
    E4_ICM = lfcShrink(my_dds, contrast=c("stage","E40_ICM","E35_ICM"), alpha=adjpvalue_cutoff, type="ashr"))
  return(res_list)
}

# extracting result table from the DESeq analysis for all comparisons
res_list <- get_results_all_comp(dds)

```

```{r ranked DE genes}

# Function which takes a DESeq results object and return it as a dataframe, with genes ordered by pvalue and filtered by adjusted pvalue
pvalue_ranked_DE <- function (my_res) {
  # keeping only genes with adjusted pvalue below my threshold
  my_res_f <- subset(my_res, padj < adjpvalue_cutoff)
  # ordering by pvalue
  my_res_f_ord <- my_res_f[order(my_res_f$pvalue),]
  my_res_f_ord$repName <- rownames(my_res_f_ord)
  return(my_res_f_ord)
}

pvalue_ranked_DE_list <- lapply(res_list, pvalue_ranked_DE)
save(pvalue_ranked_DE_list, file = "Rdata/WeiXie_development_TE/pvalue_ranked_DE_list.Rdata")
save(res_list, file = "Rdata/WeiXie_development_TE/res_list.Rdata")

```

I exclude the comparisons with the pre-EGA 2c stage because, after filtering bad samples, this stage includes only one biological replicate. 

```{r DE TEs, results='asis'}

knitr::kable(as.data.frame(pvalue_ranked_DE_list[[1]]), caption = "zygote vs MII oocyte")
knitr::kable(as.data.frame(pvalue_ranked_DE_list[[2]]), caption = "postEGA 2c vs zygote")
knitr::kable(as.data.frame(pvalue_ranked_DE_list[[3]]), caption = "4c vs postEGA 2c")
knitr::kable(as.data.frame(pvalue_ranked_DE_list[[4]]), caption = "8c vs 4c")
knitr::kable(as.data.frame(pvalue_ranked_DE_list[[5]]), caption = "ICM vs 8c")
knitr::kable(as.data.frame(pvalue_ranked_DE_list[[6]]), caption = "E3.5 ICM vs 8c")
knitr::kable(as.data.frame(pvalue_ranked_DE_list[[7]]), caption = "E3.5 TE vs 8c")
knitr::kable(as.data.frame(pvalue_ranked_DE_list[[8]]), caption = "E4.0 ICM vs 8c")

```

## MA-plots

```{r customizing plots DESeq2}

pval_hist <- function (my_dds_res) {
  ggplot(as(my_dds_res, "data.frame"), aes(x = pvalue)) +
  geom_histogram(binwidth = 0.01, fill = "Royalblue", boundary = 0) +
  ggtitle(gsub(".*: stage ", "", mcols(my_dds_res)$description[4])) +
  scale_x_continuous(breaks=c(0.0, 0.5, 1)) +
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
        title = element_text(size = 18),
        legend.text = element_text(size = 20), legend.title = element_text(size = 20),
        legend.key.height = unit(1.5,"line"))
}

MA_plot_TE <- function(my_dds_res, adjpval_to_color, l2fc_to_label, baseMean_to_label, ylims) {
  # adding label to genes which are significant and expressed above defined thresholds
  my_dds_res$to_label <- ""
  my_dds_res$to_label[!(is.na(my_dds_res$padj)) & my_dds_res$baseMean >= baseMean_to_label & my_dds_res$padj < adjpval_to_color & (my_dds_res$log2FoldChange >= l2fc_to_label | my_dds_res$log2FoldChange <= -l2fc_to_label)] <- rownames(my_dds_res[!(is.na(my_dds_res$padj)) & my_dds_res$baseMean >= baseMean_to_label & my_dds_res$padj < adjpval_to_color & (my_dds_res$log2FoldChange >= l2fc_to_label | my_dds_res$log2FoldChange <= -l2fc_to_label), ])
  my_dds_res$to_color <- ""
  my_dds_res$to_color[!(is.na(my_dds_res$padj)) & my_dds_res$padj < adjpval_to_color] <- paste0("adjpval < ", adjpval_to_color)
  dt_to_plot <- as.data.table(my_dds_res)
  dt_to_plot$repName <- rownames(my_dds_res)
  # adding color by repFamily to genes which are significant and expressed above defined thresholds
  dt_to_plot <- merge(dt_to_plot, unique(TE_anno[,.(repName,repFamily)]), by = "repName")
  dt_to_plot$is_sign <- dt_to_plot$repFamily
  dt_to_plot$is_sign[is.na(my_dds_res$padj) | dt_to_plot$padj >= adjpval_to_color] <- ""
  my.cols <- c(brewer.pal(min(12,length(unique(dt_to_plot$is_sign))), "Paired"), brewer.pal(length(unique(dt_to_plot$to_color))-12, "Pastel2"))
  my.cols[1] <- "grey"
  ggplot(dt_to_plot, aes(x = baseMean, y = log2FoldChange)) +
    geom_point(aes(color = is_sign), alpha = 0.8) +
    geom_text_repel(aes(label = to_label), size = 4, max.overlaps = 10, show.legend = FALSE) +
    scale_x_log10() +
    ylim(ylims) +
    scale_color_manual(values = my.cols) +
    ggtitle(gsub(".*: stage ", "", mcols(my_dds_res)$description[4])) +
    theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
          title = element_text(size = 20),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black"),
          panel.border = element_rect(colour = "black", fill=NA))
}

```

In the MA-plots:

* The threshold used for a dot to be coloured based on repFamily is p-value adjusted < `r adjpvalue_cutoff`.
* Transposable elements whose mean expression > `r baseMean_cutoff_to_label` and log2FoldChange > `r l2fc_cutoff_to_label` (or < -`r l2fc_cutoff_to_label`) are labeled (but labeling is limited by max.overlap parameter).

```{r MA plots, include = TRUE, fig.width=27, fig.height=24}

library(RColorBrewer)
do.call(grid.arrange, lapply(res_list[1:8], function (r) {MA_plot_TE(r, adjpval_to_color = adjpvalue_cutoff, l2fc_to_label = l2fc_cutoff_to_label, baseMean_to_label = baseMean_cutoff_to_label, ylims = c(-10,10))}))

```

## Overlap between morula-to-blastocysts' DESeq results

```{r DE TEs overlap, include = TRUE}

library(gplots)
venn(list(`ICM vs 8c` = pvalue_ranked_DE_list[[5]]$repName, `E3.5 ICM vs 8c` = pvalue_ranked_DE_list[[6]]$repName, `E3.5 tropho vs 8c` = pvalue_ranked_DE_list[[7]]$repName), show.plot=TRUE)

```

The ICM and E3.5 ICM libraries can be used almost interchangeably, while a significant fraction of DE TEs are specific of ICM or trophectoderm, thus it will be good to combine trophectoderm and ICM DESeq results for comparison with TEs diff. expr. in OgtT931del mutant blastocysts, which are whole embryo samples.

```{r sessioninfo, setup, echo = TRUE, results = 'markup'}

sessionInfo()

```